---
title: "test CellScore With Selected Hsapien Data normalized with Deseq2 normalization"
author: "Siyuan"
date: "6/29/2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, # show code
                      message = FALSE # hide console output
                      )
```
### Preparations

Before assessing data with 'CellScore', the users should first create the 
abbreviations of the cell types to be tested by 'CellScore' and a 'cell.change'
data frame.

#### Creating a map between cell types and their abbreviations
Users of CellScore should define abbreviations for the cell types they want to 
test. Each abbreviation should be unique and has a 1-to-1 mapping to a cell type.

Function 'getCellTypeMap' in this example returns a map between cell types and 
their abbreviations that are used in this tutorial. Note that this function is 
not part of the 'CellScore' library and it is only for demonstration purpose in 
this tutorial. 

```{r}
getCellTypeMap <- function() {
    ## Hardcode abbreviations for each cell type
    cellTypes <- c(
        "RPE",
        "ESC",
        "ESC.d",
        "ESC.t",
        "iPS",  ## check the comment
        "IPS",
        "URI",
        "PBMC",
        "HUVEC",

        "derived cardiomyocyte",
        "hepatocyte-like cell",
        "hepatocyte",
        "fetal liver",
        "derived dopaminergic neuron",
        "derived neuron",
        "spinal cord",
        "derived neural progenitor cell",
        "derived RPE",
        "fetal RPE",
        "fetal heart",
        "derived neural stem cell",
        "naive IPS",
        "naive ESC",
        "fetal brain",
        "derived pancreatic b-like cell",
        "derived endoderm",
        "derived foregut endoderm",
        "derived hepatic endoderm",
        "derived early hepatocyte",
        "derived late hepatocyte",
        "fibroblast",
        "induced neuron",
        "fibroblast.treated",
        "cardiomyocytes",
        "hepatocytes",
        "dopaminergic neurons",
        "neurons",
        "neual progenitor cell",
        "neural stem cells",
        "pancreatic beta cell",
        "endoderm",
        "foregut endoderm",
        "hepatic endoderm",
        "neuron",
        "lymphocyte")

    abbrv <- c(
        ## Original abbreviations
        "RPE",
        "ESC",
        "ESC.d",
        "ESC.t",
        "IPS",
        "IPS",
        "URI",
        "PBMC",
        "HUVEC",
        ## Abbreviations named by me
        "dCDM",
        "HPTL",
        "HPT",
        "fLIV",
        "dDPNEU",
        "dNEU",
        "SPC",
        "dNEUPR",
        "dRPE",
        "fRPE",
        "fHRT",
        "dNEUST",
        "nIPS",
        "nESC",
        "fBRN",
        "dPCB",
        "dEDD",
        "dFEDD",
        "dHPTEDD",
        "dEHPT",
        "dLHPT",
        "FIB",
        "INEU",
        "FIB.t",
        "CDM",
        "HPT",
        "DPNEU",
        "NEU",
        "NEUPR",
        "NEUST",
        "PCB",
        "EDD",
        "FEDD",
        "HPTEDD",
        "NEU",
        "LPC")

    cellTypeMap <- setNames(abbrv, cellTypes)
    
    # Mapping from full cell type names to their abbrvs
    print(cellTypeMap)
    cellTypeMap
}
```
#### Prepare 'cell.change' data frame
 The 'prepareSampleTransitions' prepares 'cell.change' data frame. This function 
 is not part of 'CellScore' library. function 'prepareSampleTransitions' is used 
 to generate and save the 'cell.change' in a tsv file and is not part of 
 'CellScore' library and only used in this tutorial.
 
 'cell.change' data frame contains 3 columns, which are 'start', 'test' and 'target'. 
 Each row in the data frame represents the transition in an experiment. The values 
 in the 'start' column are the abbreviations of parental cell types (or starting 
 cell types) in the transitions. The values in 'target' column are the 
 abbreviations of the target cell types in the transitions. The values in 'test' 
 column are in the format '{test_cell_type}-{start}', where '{test_cell_type}' 
 are the abbreviations of the name assigned to the test cells and '{start}' are 
 the abbreviations of parental cell types (or starting cell types). The users 
 should create 'cell.change' data frame based on their own experiments
```{r}
sampleMetadataPath <- system.file("extdata", 
                                  "sampleMetadata.csv", 
                                  package = "CellScore")

transitionPath <- system.file("extdata", 
                              "exampleTransitionsHsapienData.tsv",
                              package="CellScore")

prepareSampleTransitions <- function(sampleMetadataPath, transitionPath) {

    dat <- read.csv(sampleMetadataPath, header = TRUE)

    cellTypeMap <- getCellTypeMap()
    
    standardCellTypes <- unique(dat[dat$category == 'standard', 
                                    'general_cell_type'], 
                                drop = FALSE)

    uniqueTransitions <- unique(dat[dat$category == 'test',
                                    c('parental_cell_type',
                                      'general_cell_type')])

    colnames(uniqueTransitions) <- c("start",	"test")

    ## Exclude samples without parental cell types or without target cell type
    uniqueTransitions <- uniqueTransitions[!is.na(uniqueTransitions$start) &
                                             uniqueTransitions$start != '', ]

    ## Replace full cell type names with abbreviations
    uniqueTransitionsAbbrv <- as.data.frame(lapply(uniqueTransitions,
                                                   function(element) {
                                                      (cellTypeMap[element])
                                                   }))
    
    standardCellTypesAbbrv <- lapply(standardCellTypes, 
                                              function(element) {
                                                (cellTypeMap[element])})
    
    newUniqueTransitionsAbbrv <- data.frame(start = c(),
                                            test = c(), 
                                            target = c())
    
    for(i in 1:nrow(uniqueTransitionsAbbrv)) {
      row <- uniqueTransitionsAbbrv[i, ]
      start <- row[['start']] 
      targetTypes <- standardCellTypesAbbrv[standardCellTypesAbbrv != start]
      transitions <- sapply(targetTypes, function(target){
        c(row, target)    
        })
      transitions <- t(transitions)
      colnames(transitions) <- c('start', 'test', 'target')
      newUniqueTransitionsAbbrv <- rbind(
        newUniqueTransitionsAbbrv, transitions)
    }
    uniqueTransitionsAbbrv <- newUniqueTransitionsAbbrv

    
    ## Format 'test' column
    uniqueTransitionsAbbrv$test <- paste(uniqueTransitionsAbbrv$test,
                                         uniqueTransitionsAbbrv$start,
                                         sep = '-')

    tempDirPath <- paste(getwd(), 'temp', sep = '/')

    uniqueTransitionsAbbrv
}
```

### Calculate 'CellScore'
#### Generate 'OnOff' score by run 'OnOff' function with 'cell.change' data frame and experiment data.
First, load the dependencies and inspect the data
```{r}
# Load dependencies
library(SummarizedExperiment)
library(CellScore)
library(DEE2HsapienData)
library(dplyr)

# create 'cell.change' data frame and get the map 'CellTypeMap' between cell 
# types and their abbreviations
cell.change <- prepareSampleTransitions(sampleMetadataPath, transitionPath)
CellTypeMap <- getCellTypeMap()

# Path to the data to be assessed 
rdata.path <- system.file("extdata", "tMMNormalizedData.rds",
  package="DEE2HsapienData")
 
# Load preprocessed hsapien data in SummarizedExperiment format from DEE2
normalizedSExpr <- readRDS(rdata.path)

# replace values in 'general_cell_type' with abbrvs 
colData(normalizedSExpr)[, 'general_cell_type'] <- unname(as.array(sapply(
  normalizedSExpr$general_cell_type, 
  function(element) CellTypeMap[[element]], USE.NAMES = FALSE)))

# replace values in 'parental_cell_type' with abbrvs
colData(normalizedSExpr)[, 'parental_cell_type'] <- unname(as.array(sapply(
  normalizedSExpr$parental_cell_type, 
  function(element) {
    if (is.na(element)) {
      return(NA)
    }
    CellTypeMap[[element]]
  }, USE.NAMES = FALSE)))

# set values of 'sub_cell_type1' column
colData(normalizedSExpr)$sub_cell_type1 <- paste(
  colData(normalizedSExpr)$general_cell_type, 
  colData(normalizedSExpr)$parental_cell_type,
  sep = '-')

# Set feature_id for of 'rowData' of the input
sRRAccessions <- rownames(assay(normalizedSExpr, 'counts'))
rowData(normalizedSExpr) <- matrix(sRRAccessions, 
                                   , 
                                   ncol = 1, 
                                   dimnames = list(sRRAccessions, c('feature_id')))

# Set assay 'exprs' same as 'counts'
assays(normalizedSExpr)$exprs <- assay(normalizedSExpr, "counts")

coldata <- colData(normalizedSExpr)  
# Set 'donor_tissue' as 'parental_cell_type'
colData(normalizedSExpr)$donor_tissue <- coldata$parental_cell_type
# Set the values of 'experiment_id' as values in 'SRX_accession' column of
# the coldata of the input object
sampleMetaData <- read.csv(sampleMetadataPath, header = TRUE)
# Set the values in 'platform_id' as values in 'Platform' column of the
# coldata of the input object.
colData(normalizedSExpr)$platform_id <- colData(normalizedSExpr)$Platform
# Set the values in 'sample_id' as SRR accessions, which are the rownames
# of the coldata of the input object.
colData(normalizedSExpr)$sample_id <- rownames(colData(normalizedSExpr))
# Set the values in 'experiment_id'.
colData(normalizedSExpr)$experiment_id <- colData(normalizedSExpr)$SRX_accession

PcaPlot(normalizedSExpr)

```
```{r}
group.OnOff <- OnOff(normalizedSExpr, cell.change, out.put="marker.list")
print(summary(group.OnOff))
```

```{r}
individ.OnOff <- OnOff(normalizedSExpr, cell.change, out.put="individual")
```

```{r}
barplot.out <- BarplotOnOff(normalizedSExpr, group.OnOff$scores)
barplot.out
```

```{r}
tmp.time <- system.time(cs <- CosineSimScore(normalizedSExpr, cell.change,
                                             iqr.cutoff=0.1))
PlotCosineSimHeatmap(cs$cosine.general.groups, "general groups",
                     width=20, height=20, x=-20, y=3)

```

```{r}
cellscore <- CellScore(normalizedSExpr, cell.change, individ.OnOff$scores,
                       cs$cosine.samples)

```

```{r}
ScatterplotCellScoreComponents(cellscore, cell.change, FALSE)

```

```{r}
pdf(file="DESeq2 Normalized Data CellScore Report (all transitions).pdf",
    width = 8, height = 9)
CellScoreReport(cellscore, cell.change, group.OnOff$markers, normalizedSExpr,
                cs,
                group.by = 'transition')
dev.off()
```
